Basic Model formulation

### 2015-01-29 documentation/model_formulation.Rmd
update model formulation, include ODE and CA models

G <- function(P, r = 2, K = 0.9) r*P*(1-P/K)

@@

The logistic growth model is phenomenological and does not provide a particular ecological explanation, why growth is decelerating to zero with increasing cover. It assumes just some form of global competition leading to a balance of gains and losses at the carrying capacity of the ecosystem.

@@

@@

These are global feed-backs. But local feedbacks also could lead to this deceleration:  
For instance, if plant cover would have a dominantly negative effect on it's surroundings, by depleting nutrients or light, the local neighborhood of a plant cannot be occupied as easy as more remote places. As cover increases and the interspace areas are closing in, the space available for rejuvenation approaches zero.

@@

In arid ecosystems, growth would be low at very low cover and accelerate as cover increases, because retention of water and organic matter in the site is fundamentally reduced. This postive feed-back (low cover reduces growth, leading to even lower cover) was discussed by @vandekoppel97 and @rietkerk97 to explain bistability in arid ecosystems.

@@

$$ r = r_0 * V^a $$

$$ G(V) = r_0 V^{a+1} (1-\frac{V}{K}) \,,$$

@@

r <- function(P, r_0 = 2, a = 2) r_0*P^a

defplot(rho, r(rho), ylab = "growth")

G_sig <- function(P, r_0 = 2, a = 2, K = 0.9) r(P, r_0, a)*P*(1-P/K)

defplot(rho, G_sig(rho), ylab = "growth")

### functional response

@@

In the cellular automata model and mean-field approximations of @kefi07, individual plant mortality is a constant rate, which means it is increasing linearly with vegetation cover.

Plant mortality due to grazing therefore would be described as

@@

thus would be defined as constant, $g$, per grazer individual regardless of the vegetation cover.

@@

In natural systems, however, consumption is defined by saturating functional responses, i.e. at low cover, density dependent mortality defined by search efficiency dominates, but at high cover handling times of the consumer will limit its intake rates.

Therefore, plant mortality due to grazing should be

@@

Since we are calculating in fractions of landscape, and not in cells, the handling time defines the time a grazer individual would need to handle the entire landscape as readily available fodder. The search time is the percentage of the landscape foraged in over the course of one year.

@@

@@

We implement the accelerating and decelerating features on a local level.

@@

This is assuming that growth rate is enhanced by a local facilitation term similar to the implementation in Kéfi et al 2007. The growth rate becomes a function of local vegetation cover $\nu$:

@@

This is assuming that growth is diminished locally by competition, $c$. This affects carrying capacity rather than growth rate itself.

@@

$$ $$

>>>>

G <- function(P, r = 2, K = 1) r*P*(1-P/K)

@@

In the first place the logistic growth model is phenomenological and does not provide a particular ecological explanation, why growth is decelerating to zero with increasing cover. It assumes just some form of global competition leading to a balance of gains and losses at the carrying capacity of the ecosystem.

@@

We consider these multiple competitive effects being of minor relevance for now and start with a single limitation by the factor space. As the landscape becomes more and more filled with vegetation, the empty space become scarce. Thus, growth declines until each potentially habitable spot is occupied. We implement a generally valid carrying capacity, $K$.

@@

We assume that growth rate is linearly related to environmental quality, $b$, which is inverse to aridity. Since drylands are water limited in the first place, this factor defines the reproduction rate (i.e. number of seeds produced per plant, number of seedlings surviving the harshness of the dry soils). We assume that

$$ r = r_0 * b $$

to have a gradient of environmental harshness, $b$, reaching from 0 to 1. 
The model thus starts out with the assumption of decelerating growth due to competition for space.

@@

In arid ecosystems, growth rates would be lowest at very low cover and accelerate as cover increases, because retention of water and organic matter in the site is fundamentally reduced and water runoff is high. This postive feed-back (low cover reduces growth, leading to even lower cover) was discussed by @vandekoppel97 and @rietkerk97 to explain bistability in arid ecosystems.

@@

$$ r = r_0 b V^a $$

$$ G(V) = r_0 b V^{a+1} (1-\frac{V}{K}) \,,$$

@@

r <- function(P, r_0 =  2.25, a = 1) r_0 *P^a

#defplot(rho, r(rho), ylab = "growth")

G_sig <- function(P, r_0 = 2.25, a = 1,  K = 0.9) r(P, r_0, a)*P*(1-P/K)

defplot(rho, G_sig(rho), ylab = "growth")
lines(rho, G_sig(rho, a = 0))
lines(rho, G_sig(rho,r_0 = 2.25, a = 1))
lines(rho, G_sig(rho,r_0 = 2.25, a = 0.2))
```

### plant mortality

@@

In the cellular automata model and mean-field approximations of @kefi07, individual plant mortality is a constant rate, which means it is increasing linearly with vegetation cover. Total plant mortality due to grazing therefore would be described as

@@

thus could be defined as constant, $g$, per grazer individual regardless of the vegetation cover.

@@

In natural systems, however, consumption is defined by saturating functional responses, i.e. at low cover, density dependent mortality defined by search efficiency dominates, but at high cover handling times of the consumer will limit its intake rates. Therefore, plant mortality due to grazing should be

@@

Since we are calculating in fractions of landscape, i.e. vegetation cover, and not in cells, the handling time defines the time a grazer individual would need to handle the entire landscape as readily available fodder. The search time is the percentage of the landscape foraged in over the course of one year.

@@

### population dynamics

The differential equation describing population cover above with $V$ and $C$ substituted would then become

$$ \frac{\mathrm{d}\rho_1}{\mathrm{d}t} = r b V (1-\frac{V}{K}) - m V - \frac{aVL}{1+ahV} $$

Given a parameter set  
```{r, echo=FALSE}

parms <- list(
    m = 0.05,
    r = 1,
    b = 0.9, 
    K = 0.9, 
    a = 0.6,
    h = 100,
    L = 10
  )

 as.data.frame(parms)
```
we receive a projection of the intersection of growth and mortality.

```{r, echo=FALSE}
plot(rho,  with(parms,  m * rho + (a * rho * L)/(1 + a * h * rho)), ylab = "growth", type = "l", ylim = c(0,0.5))
lines(rho, with(parms, r* b* rho * (1 - rho/K) ), col = "green3" )

And solving the ODE, starting with values larger than the unstable equilibrium point on the left,  returns a timeseries

```{r, echo=FALSE}
d_rho <- function(rho, parms, z = 4) { 
  with(parms, r* b* rho * (1 - rho/K) - m * rho - (a * rho * L)/(1 + a * h * rho) )  
}

odesys <- function (t, rho, parms = model_parms) {
  list( d_rho(rho, parms)  )
}

library(deSolve)
# running the ode-solver
runmodel <- ode(y = 0.90, func = odesys, times = 1.1^(1:53), parms = parms)

# transfer into ouput and calculate missing rho values
ODEnullmodel <- as.data.frame(runmodel)

plot(ODEnullmodel, type = "l", ylim = c(0,1))

### Cellular Automata

The differential equations are formulated as changes in vegetation cover. Growth are positive changes, and deaths are negative changes.

Now, the CA uses transition probabilities. That is, a) the individual risk of death of vegetated cells and b) the individual chance of becoming vegetated for an empty cell.
To translate into the CA context, we need to find rules that apply to a) the vegetated cells, and b) the empty cells. Thus, it would be a) the reduction of term of the population of vegetated cells, divided by cover, to get the individual risk of death

$$ w_{1,0} = m + \frac{aL}{1+ah\rho_1} $$

This would be b) the growth term for the population of empty cells, divided by cover of empty cells, to get the individual chance of growth.

$$ w_{0,1} = \frac{r \rho_1 b (1-\frac{\rho_1}{K})}{1-\rho_1} $$

From an individual cell perspective, the probabilities would change along with cover like this

```{r, echo=FALSE}
par(mar = c(4,4,2,4))
plot(rho,  with(parms,  r * b *  ( 1 - (rho / K)) / (1 - rho) ) , ylab = "individual growth on empty cells", type = "l", ylim = c(0,2.5), las = 1, yaxs = "i")

lines(rho, with(parms, m + ( a * L )/( 1 + a * h * rho ) ), col = "red3" )
axis(4, las = 1)
mtext("individual risk of death of vegetated cells", 4, 3)
```

Running this in a timeseries of 150 years yields

```{r, echo = FALSE}

source("../code/simfunctions.r")

set.seed(584232) # setting seed for random number generation

nullmodel <- runCA(0.9, parms, delta = 0.2, t_max = 150, t_min = 100, t_eval = 50, isstable = 0.0001, saveeach = 5)

plot(nullmodel$rho_one ~ nullmodel$time, 
     ylim = c(0,1), ylab = expression( rho["+"]), 
     xlab = "time [yr]", type = "l")

lines(ODEnullmodel, lty = 3)

#animateCA(nullmodel, "ca_nullmodel.gif")

@@

We implement accelerating and decelerating features on growth and mortality on the local level.

@@

This is assuming that the local environmental suitability is enhanced by a local facilitation term similar to the implementation in Kéfi et al 2007. Thus, the growth rate becomes a function of local vegetation cover $\nu$:

@@

For instance, if plant cover would have a dominantly negative effect on it's surroundings, by depleting nutrients or light, the local neighborhood of a plant cannot be occupied as easy as more remote places. As cover increases and the interspace areas are closing in, the space available for rejuvenation approaches zero. 
This is assuming that growth is diminished locally by competition, $c$. This affects carrying capacity rather than growth rate itself.

@@

### implementation in a cellular automaton

It is straigthforward to implement the model in a cellular automata model. It is defined on a defined grid of cells where each cell can take the values '1' if it is vegetated and '0' if it is empty. The transition rules between those states depend either on global vegetation cover $\rho_1$ or on the local cover in the direct neighborhood of the cell $\nu_1$ (closest 4 cells, von-Neumann-neighborhood).

Thus, two transition probabilities need to be defined: the probability of **growth**, $w_{0,1}$, turning an empty cell into a vegetated, and the probability of **death**, $w_{0,1}$, removing vegetation from the cell.

The null model, i.e. the model with only global effects on mortality and growth, with the basic assumption of a type II functional response and logistic growth, would thus be defined by

$$ w_{0,1} = r (1-\frac{\rho_1}{K})$$

$$ w_{1,0} =  m + \frac{aL}{1+ah\rho_1} $$

### implementation in a pair-approximation model

The pair-approximation framework (Matsuda et al 1992) is defined as a system of ordinary differential equations, describing the populations of *pairs* of cells. In a two state system, three pairs can be defined with population densities $\rho_{1,1}$ , $\rho_{0,0}$, and $\rho_{0,1}$. Besides, there are the population densities of singletons, $\rho_0$ and $\rho_1$.

There are conservation equations which allow to calculate population densities of one pair out of the other pairs and singletons.

$$ \rho_1 + \rho_0 = 1$$

$$ \rho_{1,1} + \rho_{0,1} = \rho_1$$

$$ \rho_{0,1} + \rho_{0,0} = \rho_0$$

This means, the system can be fully described by two differential equations:

$$ \frac{\mathrm{d}\rho_{1}}{\mathrm{d}t} = $$

$$ \frac{\mathrm{d}\rho_0}{\mathrm{d}t} = $$
